---
output: github_document
editor_options: 
  chunk_output_type: console
bibliography: "references.bib"
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}

ratio <- (1 + sqrt(5)) / 2

knitr::opts_chunk$set(
  # cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  screenshot.opts = list(
      vwidth = floor(800 * ratio),
      vheight = 800
  ))
```

# Plotscaper

<!-- badges: start -->
<!-- badges: end -->

Plotscaper is an R package designed for making interactive figures geared towards data exploration. All plots in `plotscaper` figures support linked selection, as well as wide variety of other interactions, including parameter manipulation, zooming, panning, reordering, and others.

## Quick start

To get started, install `plotscaper` with:

```{r}
#| eval: false
devtools::install_github("bartonicek/plotscaper")
```

Next, open up RStudio and run the following code:

```{r}
library(plotscaper)

layout <- matrix(c(
  1, 1, 2, 3,
  1, 1, 4, 5,
  6, 7, 7, 7
), ncol = 4, byrow = TRUE)

set_scene(sacramento) |>
  add_scatterplot(c("longitude", "latitude")) |>
  add_barplot(c("city")) |>
  add_histogram(c("sqft")) |>
  add_fluctplot(c("beds", "baths")) |>
  add_histogram2d(c("sqft", "price")) |>
  add_notes() |>
  add_parcoords(names(sacramento)) |>
  set_layout(layout)
```

In your viewer, you should now see something like the image above, however, your version should be fully interactive (Github doesn't allow JavaScript in `README.md`, hence why the image above is static, however, the other vignettes should work okay). 

Try clicking and dragging to select a few points on the scatterplot!
