---
output: github_document
editor_options: 
  chunk_output_type: console
bibliography: "references.bib"
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}

ratio <- (1 + sqrt(5)) / 2

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/readme1.png",
  screenshot.opts = list(
      vwidth = floor(800 * ratio),
      vheight = 800
  ))
```

# Plotscaper

<!-- badges: start -->
<!-- badges: end -->

Plotscaper is an R package designed for making interactive figures for data exploration. All plots in `plotscaper` support linked selection by default, as well as wide variety of other interactions, including, zooming, panning, reordering, and parameter manipulation.

The package powering `plotscaper` figures ([plotscape](https://github.com/bartonicek/plotscape)) is written in (mostly) vanilla TypeScript/JavaScript and uses no external dependencies/framework for reactivity.

## Quick start

To get started, install `plotscaper` with:

```{r}
#| eval: false
devtools::install_github("bartonicek/plotscaper")
```

Next, open up RStudio and run the following code:

```{r}
library(plotscaper)

names(airquality) <- c("ozone", "solar radiation", "wind", 
                       "temperature", "month", "day")

create_schema(airquality) |>
  add_scatterplot(c("solar radiation", "ozone")) |>
  add_barplot(c("day", "ozone"), list(reducer = "max")) |>
  add_histogram(c("wind")) |>
  add_pcoords(names(airquality)[1:4]) |>
  render()
```

In your viewer, you should now see something like the (static) image above, however, your version should be fully interactive. Alternatively, to see a fully interactive version of the figure online, go to [Getting started](plotscaper.html) (the image above is just a static snapshot because Github doesn't allow JavaScript in `README.md`. The other vignettes are full HTML files rendered with RMarkdown and do support JavaScript as a result).

Try clicking and dragging the scatterplot to select a few points!
