---
title: "Figure layout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Figure layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

We can control the figure layout by using the `set_layout` function. This function works similarly to the `layout` function from the `graphics` package. We just give the function a matrix of numeric values representing the plot ids, and the figure automatically resizes the plots based on how many equal-sized rectangles in a grid each plot takes up. 

For example, here's how we can create a figure with large scatterplot on the top-left, a tall histogram on the right-hand side, a short wide histogram on the bottom, and a small section for notes (not-(e)-plot, pardon the pun) on the bottom-right:

```{r}
#| fig-width: 7
#| fig-height: 7

library(plotscaper)

layout <- matrix(c(
  1, 1, 2,
  1, 1, 2,
  3, 3, 4
), ncol = 3, byrow = TRUE)

set_scene(sacramento) |>
  add_scatterplot(c("sqft", "price")) |>
  add_histogram("price") |>
  add_histogram("sqft") |>
  add_notes() |>
  set_layout(layout)

```

You can still resize individual plots by pressing and holding the `S` key and then dragging a widget in the bottom right of the plot area. Note, however, that this feature is a bit finicky and may result in gaps. 
