---
title: "Getting Started with Plotscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Plotscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Getting started with `plotscape` is easy! Use `ps_scene(data, width, height)` function to create a frame to hold all your interactive plots, of pre-specified dimensions, and finally add as many interactive plots as you want with `ps_[plottype]`. For example, here's how you make an interactive scene with the famous `mtcars` data:

```{r setup}

library(plotscapeTest)

ps_scene(mtcars, width = 600, height = 400) |>
  ps_scatter(ps_map("wt", "mpg")) |>
  ps_bar(ps_map("cyl")) |>
  ps_histo(ps_map("disp")) |>
  ps_square(ps_map("gear", "am"))

```

Once you have a scene set up, you can interact with it by using mouse/touchpad and pressing keys on a keyboard. Some useful commands are:

- Click a plot to make it active
- +/- to increase/decrease size of objects in the (active) plot
- [/] to decrease/increase opacity of objects
- Click an object to transiently select it, click-and-drag to select more than one object
- Hold down 1, 2, ... while clicking (or clicking-and-dragging) to make a permanent (group) selection

It's easy to use `plotscape` programmatically. For example, here's how you can set up a scatterplot-matrix (SPLOM) of the `iris` dataset:

```{r}

# Create a matrix of all unique combinations of 2 variables in the iris data
vars <- names(iris)
xy <- matrix(vars[combn(1:4, 2)], ncol = 2, byrow = TRUE)

lay1 <- matrix(c(
  1, 7, 7, 2, 
  3, 7, 7, 4,
  5, 7, 7, 6
), ncol = 4, byrow = TRUE)

s1 <- ps_scene(iris, layout = lay1, width = 600, height = 400)
for (i in seq_len(nrow(xy))) {
  s1 <- s1 |> ps_scatter(ps_map(xy[i, 1], xy[i, 2]))
}
s1 <- s1 |> ps_bar(ps_map("Species"))
s1

```
