---
title: "Layout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Adding and removing plots

As you have probably seen in other vignettes, you can add plots using the various `add_plot` functions:

```{r}

library(plotscaper)

names(airquality) <- c("ozone", "solar radiation", "wind", 
                       "temperature", "month", "day")

schema <- create_schema(airquality) |>
  add_scatterplot(c("solar radiation", "ozone")) |>
  add_barplot(c("day")) |>
  add_histogram(c("wind")) |>
  add_pcoords(names(airquality))

schema |> render()

```

You can also remove plots:

```{r}
schema |> 
  remove_plot("bar1") |> # Remove the first barplot added to the figure
  pop_plot() |> # Remove the last plot added to the figure - typically, the bottom-right one
  render()
```

This does not make much sense server-side, since, if you don't want a specific plot to be in the figure, you can just not add the figure in the first place. However, on the client-side, this can be very useful, since you can modify the figure on the fly:

```{r}
#| eval: false
# NOT RUN - this only makes sense inside a running R session
scene <- create_schema(airquality) |> render()

scene |> add_scatterplot(c("solar radiation", "ozone"))
scene |> add_barplot(c("day")) # Oops, maybe I want a histogram instead
scene |> pop_plot()
scene |> add_histogram(c("wind")) 
```

Some of this was already discussed in the [Get started](plotscaper.html) section.

## Layout

We can also modify the layout of a plotscaper figure, using an interface that's very similar to `graphics::layout`. If you're not familiar with this function, we can provide it a numeric matrix, with values identifying plotting regions as contiguous rectangles, and the plots will be appropriately resized:

```{r}

# 1. Big scatterplot on the top left
# 2. Small barplot on the top right
# 3. Small histogram on the right
# 4. Wide parallel coordinates plot on the bottom
# 0. (empty space on the bottom right)

layout <- matrix(c(
  1, 1, 2,
  1, 1, 3,
  4, 4, 0
), ncol = 3, byrow = TRUE)

schema |> set_layout(layout) |> render()
```

The matrix can be of any size, the only requirement is that the plot areas must be contiguous, meaning that e.g. this would not be a valid layout:

```{r}
#| eval: false

layout <- matrix(c(
  2, 1, # Cannot split plotting regions like this
  1, 2
), ncol = 2) 
```

Also, all of the plots should be included in the layout. If they are not, a pop-up warning will show up. You can decide whether you want to proceed with the figure, however, be warned that the layout may not work well.

Finally, you can call the `set_layout` function from inside a running R session with `scene |> set_layout(...)` (and you will get a client-side update to the current figure).
